---
title: "Using rdz"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

rdz writes R objects to files (or raw vectors) that can be read back whole or
in part. This article walks through a typical session: writing and reading,
reading only what you need, looking inside a file without reading it, caching
on content hashes, and keeping several objects in one file.

```{r setup}
library(rdz)
```

## Writing and reading

`write_rdz()` and `read_rdz()` work as `saveRDS()` and `readRDS()` do.

```{r}
path <- tempfile(fileext = ".rdz")
write_rdz(mtcars, path)
identical(read_rdz(path), mtcars)
```

Logical, integer, double and character vectors, factors, lists and data
frames, nested and with their attributes, are stored *natively*: column by
column, in checksummed blocks, each vector with the encoding that suits it.
Anything else (an environment, a function, a model with a formula) goes
through R's own serializer in the same blocks. Either way the value comes back
as written.

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
write_rdz(fit, path)
rdz_info(path)$codec
all.equal(read_rdz(path), fit)
```

`mode = "native"` makes `write_rdz()` fail instead of falling back, for code
that relies on selective reads, and `mode = "r"` always uses R's serializer.

## Presets and threads

Two options choose how blocks are stored, and neither changes what a reader
needs: any setting reads any file.

```{r}
set.seed(1)
sales <- data.frame(
  day = as.Date("2026-01-01") + 0:99999 %/% 100,
  store = factor(sample(c("north", "south", "east", "west"), 1e5, TRUE)),
  units = sample.int(50L, 1e5, TRUE),
  price = round(runif(1e5, 1, 100), 2)
)

sizes <- sapply(c("speed", "balanced", "compact"), function(preset) {
  old <- options(rdz.preset = preset)
  on.exit(options(old))
  write_rdz(sales, path)
  file.size(path)
})
rds <- tempfile(fileext = ".rds")
saveRDS(sales, rds)
c(sizes, saveRDS = file.size(rds))
```

`"balanced"` (the default) compresses each block with Zstandard at level 1;
`"compact"` uses a higher level; `"speed"` stores blocks uncompressed.
`options(rdz.threads = 4)` compresses and decompresses blocks on four
threads, and the file is byte for byte the same whatever the number.

## Reading only what you need

`select` takes columns of a data frame (or elements of a list) by name or
position, and `rows` takes rows of a data frame (or elements of a vector).
Only the blocks that hold them are read and decoded.

```{r}
write_rdz(sales, path)
read_rdz(path, select = c("day", "price"), rows = 1:5)
read_rdz(path, rows = c(99999, 100000))
```

The result is what `[` gives on the whole value, so classes, factor levels and
automatic row names come through as they would in memory.

## Looking inside a file

`rdz_info()` reads the header and directory, never the data. `rdz_schema()`
lists a native file's tree, with each part's class, shape, attributes and
stored size.

```{r}
report <- list(sales = sales, note = "Q1", models = list(fit = coef(fit)))
write_rdz(report, path)
rdz_schema(path)
```

`rdz_attributes()` reads the attributes of any part, given its path through
the tree as names or positions, without decoding that part's data.

```{r}
rdz_attributes(path, object = c("sales", "store"))
rdz_attributes(path, object = c("models", "fit"))
```

Files can also carry their own key-value metadata, such as a source or a code
version, which `rdz_info()` returns without reading the value.

```{r}
write_rdz(sales, path, metadata = c(source = "pos-export", version = "2026.10"))
rdz_info(path)$metadata
```

## Caching on content

Every file records a 128-bit hash of the value it holds, and `rdz_hash()`
gives the same hash for an object in memory, whatever the preset, the threads
or the platform. That makes a file its own cache key.

```{r}
rdz_info(path)$content_hash
rdz_hash(sales)
```

With `skip_unchanged = TRUE`, `write_rdz()` leaves a file that already holds
the value untouched, bytes and modification time alike, so build tools and
caches keyed on files see no change.

```{r}
before <- file.mtime(path)
Sys.sleep(1)
write_rdz(sales, path, skip_unchanged = TRUE,
          metadata = c(source = "pos-export", version = "2026.10"))
identical(file.mtime(path), before)
```

`rdz_verify()` checks every block's checksum without building the value, and
with `content = TRUE` it also checks that the value hashes to the stored hash.
It returns the path, invisibly, when the file is sound, and an error says
which check failed when it is not.

```{r, error = TRUE}
rdz_verify(path, content = TRUE)

damaged <- readBin(path, "raw", file.size(path))
damaged[200] <- xor(damaged[200], as.raw(0xff)) # one flipped byte
writeBin(damaged, path)
rdz_verify(path)
```

## In memory

`rdz_serialize()` gives the file's bytes as a raw vector, for a database
column, a message queue or a socket, and every reader takes a raw vector where
it takes a path.

```{r}
bytes <- rdz_serialize(sales)
length(bytes)
read_rdz(bytes, select = "units", rows = 1:3)
identical(rdz_unserialize(bytes), sales)
```

## Several objects in one file

`rdz_save()` and `rdz_load()` work as `save()` and `load()` do, and
`rdz_load()` reads only the objects you name.

```{r}
archive <- tempfile(fileext = ".rdz")
rdz_save(sales, fit, file = archive)
rm(sales, fit)
rdz_load(archive, names = "fit")
exists("fit")
exists("sales")
```

The archive is an ordinary rdz file holding a named list, so `read_rdz()` and
`rdz_schema()` work on it too.

```{r, include = FALSE}
unlink(c(path, archive, rds))
```
